---
title: "Data Wrangling in tidyr"
author: "Javier Patron"
date: "2022-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

library(tidyverse)
library(here)
library(janitor)
```


## Read in the world bank data files


```{r}
wb_indicators <- read_csv(here::here("data", "wb_indicators.csv"), na = c("..", ""))
world_metadata <- read_csv(here::here("data", "wb_indicators_metadata.csv"))
```

### Tidy up the data

QUESTION: ```{r} can you add names to filter the columns ?
wb_indicators_long <- wb_indicators |> 
  pivot_longer(cols = names(-"Country Name", "Country Code", "Series Name", "Series Code", names_to = "years", values_to = "indicator_value"))

```{r}
wb_indicators_long <- wb_indicators |> 
  pivot_longer(cols = "2001 [YR2001]": "2020 [YR2020]", 
               names_to = "years", 
               values_to = "indicator_value")
```


## Separate to clean up the year column

```{r}
wb_clean <- wb_indicators_long |> 
  separate(col = years, into = c("year", "year_chr"), sep = " ") |> 
  select(-year_chr, -"Country Code", -"Series Code") |> 
  mutate(year = as.numeric(year))|> 
  drop_na("Series Name") |> 
  pivot_wider(names_from = "Series Name", values_from = indicator_value)


```

### Rename the titles for each column
```{r}

names(wb_clean) <- c("country", "year", "access_clean_fuels_pp", "access_electricity_pp", "co2_emissions_kt", "fossil_fuel_cons_pct", "water_specs")
```


##Subset for specific countries

```{r}
wb_subset <- wb_clean |> 
  filter(country %in% c("Algeria", "Barbados","Bulgaria", "Chile"))
```

## Lets Graph!

```{r}

ggplot(data = wb_subset, aes(x = year, y = co2_emissions_kt)) +
  geom_line(aes(color = country)) +
  facet_wrap(~country)
```


### Filter and get data

## Example 1: Filter to only keep exact matches

```{r}
ex_1 <- starwars |> 
  filter(height > 180)
```

### Get the exact number match

```{r}
ex_2 <- starwars |> 
  filter(eye_color == "blue")
```

### Example 3: Make a subset of starwars for only NABU as the home world

```{r}
ex_3 <- starwars |> 
  subset(homeworld == "Naboo")
```



### Double conditions in a data frame

```{r}

ex_4 <- starwars |> 
  filter(height > 180 & homeworld == "Tatooine")
  
```

### Example5: Create a subset where the hair color is brown and the species is human

```{r}


ex_5 <- starwars |> 
  filter(hair_color == "brown" & species == "Human")
```

### Example 6: Filtering OR statements

```{r}
ex_6 <- starwars |> 
  filter(height > 180 | eye_color == "yellow")

```

### Example 7: Create a subset that keep rows if homeworld is Endor OR species is droid

```{r}
ex_7 <- starwars |> 
  filter(homeworld == "Endor" | species == "droid")
```


### Example 8: Eye color (blue,brown,red)

```{r}

ex_8 <- starwars |> 
  filter(eye_color == "blue" | eye_color == "red" | eye_color == "brown")


ex_9 <- starwars |> 
  filter(eye_color %in% c("brown", "red", "blue"))


# YOU DONT WANT TO DO THIS

ex_10 <- starwars |> 
  filter(eye_color == c("brown", "red", "blue"))

```

### Example 11. Create a subset where homewoeld is endor or Tattoine or naboo and height < 150

```{r}
ex_11 <- starwars |> 
  filter(height < 150, homeworld %in% c("Endor", "Tatooine", "Naboo"))

ex_12 <- starwars |> 
  filter(species != "Human" )
```

### Use the subset

```{r}
ex_13 <- subset(starwars, species == "Human", select = c(name, sex, gender,species))
```

## Use `select`: select or exclude columns

```{r}

# Select by name

ex_14 <- wb_clean |> 
  select(country,year, co2_emissions_kt)

ex_15 <- wb_clean |> 
  subset(select = c(country, year, co2_emissions_kt))


```


Multiple pipes together

```{r}

ex_17 <- wb_clean |> 
  filter(country %in% c("Bolivia", "Chile", "Mexico")) |> 
  select(country:fossil_fuel_cons_pct)

ex_18 <- subset(wb_clean, country %in% c("Bolivia", "Chile", "Mexico"), select = country:fossil_fuel_cons_pct)
```



Example 19- SELECT ONLY COLUMNS FOR COUNTRY, YEAR, AND WATER STRESS AND REORDER THEM AS YEAR, COUNTRY WATER STRESS, THEN FILTER TO ONLY INCLUDE OBSERVATIONS FOR ANY THREE COUNTRIES OF YOUR CHOOSING

```{r}

ex_19 <- wb_clean |> 
  select(country, year, water_specs) |> 
  filter(country %in% c("Fiji"))

```


### Change the names of the columns `rename`

```{r}
ex_20 <- wb_clean |> 
  rename(ffcp = fossil_fuel_cons_pct, ws = water_specs)

```

`mutate`: update an existing column, or add a new one
If you assign a NEW column name, it will add a new column to the end of your data frame. If you assign an existing column name, it will overwrite that column

```{r}

ex_22 <- wb_clean |> 
  mutate(co2_emissions_t = co2_emissions_kt * 1000)

ex_23 <- wb_clean |> 
  mutate(co2_emissions_t = co2_emissions_kt * 1000) |> 
           relocate(co2_emissions_t, .after = co2_emissions_kt)

ex_24 <- wb_clean |> 
  mutate(year = as.character(year))

ex_25 <- wb_clean |> 
  mutate(year = as.numeric (year))
```


### dplyr relocate the columns

```{r}

ex_26 <- wb_clean |> 
  relocate(water_specs, .after = year)
```

### Use of group_by

group_by: Creates groups under the hood . Doesnt chance anything superficial. Under the hood os where summarize comes in
summarize: apply youir functions to EACH GRPUP, and return them in a nice summary table

```{r}

# Calculate the average height by species

ex_27 <- starwars |> 
  filter(homeworld %in% c("Naboo")) |> 
  group_by(species) |> 
  summarize(mean_height = mean(height, ra.rm = TRUE),
            mean_mass = mean(mass, ra.rm = TRUE))
            
ex_28 <- starwars |> 
  group_by(species) |> 
  summarise(mean_height = mean(height, ra.rm = TRUE),
            mean_mass = mean(mass, ra.rm = TRUE))

```


```{r}

df <- tribble(
  ~species, ~length, ~number,
  "lobster", 12, 4,
  "lobster", 14, 6,
  "lobster", 15, 2,
)

df_case <- df |> 
  uncount(number)

```



### count function for observations
# combination of group_by summarize n()

```{r}

starwars |> 
  count(homeworld, species)

starwars |> 
  group_by(homeworld,species) |> 
  summarize(size = n(),
            mean_height = mean(height, na.rm= TRUE))
```


Practice example:
- Start from wb_clean
- Only keep observations for Nicaragua, Ecuador, peru and Chile, AND for years greater than 2005
- Only co2 emissions
- find the toal co2 emissions converted to megatons


```{r}
ex_29 <- wb_clean |> 
  filter(country %in% c("Nicaragua", "Ecuador", "Peru", "Chile") & year > 2005) |>
  select(country, year, co2_emissions_kt) |> 
  mutate(co2_emissions_kt, co2_emissions_MT = co2_emissions_kt/1000) |> 
  group_by(country) |> 
  summarize(total_co2_mt = sum(co2_emissions_MT, na.rm = TRUE))
  
```
